---
title: "CaseStudy4"
output: html_notebook
author: "Dustin Bracy, Grace Lang, Paul Huggins, Branum Stephan"
date: "1/24/2020"
editor_options: 
  chunk_output_type: console
---

```{r library install}
library(tidyverse)
library(DataExplorer)
library(dplyr)
library(ggplot2)
library(esquisse)
library(rvest)
library(stringr)
library(chron)
library(data.table)
library(scales)
```


# Data Extraction, Transformation & Loading:

## Extraction via web scraping
```{r Web scraper function, eval=FALSE}
# parse page-wise data for a given year
parse_page <- function(year){
  output = c()
  page = 1
  while(TRUE){
    run_results <- read_html(paste("http://www.cballtimeresults.org/performances?division=Overall+Women&page=", page, "&section=10M&sex=W&utf8=%E2%9C%93&year=", year, sep=""))
    results <- data.frame(run_results %>% html_table())
    if(nrow(results) > 0) {
      output = rbind(output, results)
      print(paste("Processing for page: ", page))
      page = page + 1
    }
    else{
      break
    }
  }
  return(output)
}

# loop through multiple years
pull_all_run_info <- function(){
  total_output = c()
  for(year in c(1999: 2012)){
    print(paste("\tfetching for year:", year))
    total_output = rbind(total_output, parse_page(year))
  }
  return(total_output)
}

# run scraper
cb_data <- pull_all_run_info()

```

## Data Transformation
```{r save copies and write to csv for record, eval=FALSE}
# save original copy of data so we don't have to re run scraper
cb_orig <- cb_data

# write out original to csv for record
# write.csv(cb_orig, "/Unit4_CaseStudy/data/CB_Raw_Data.csv", header = TRUE)

# save copy to manipulate and parse
cb_race <- cb_data
```

```{r data cleanup and feature engineering, eval=FALSE}
cb_race <- as.data.frame(cb_race)

# drop 10M from the first column and then rename Name to Year
cb_race$Race <- str_split(cb_race$Race, " ", simplify = TRUE)
colnames(cb_race)[1] <- "Year"

# drop (W) from the second column since all participants are Women
cb_race$Name <- gsub("\\s*\\([^\\)]+\\)","",as.character(cb_race$Name))

# split the hometown column into each part
hometownsplit <- data.frame(do.call("rbind", strsplit(as.character(cb_race$Hometown), ",", fixed = TRUE)))
hometownsplit$X3 <- ifelse(hometownsplit$X1 %in% hometownsplit$X2, hometownsplit$X2, "NaN")
hometownsplit$X1 <- ifelse(hometownsplit$X1 %in% hometownsplit$X2, "NaN", hometownsplit$X1)
hometownsplit$X3 <- ifelse(hometownsplit$X3 %in% c("NaN"), "USA", hometownsplit$X3)
hometownsplit$X2 <- gsub('\\s+', '', hometownsplit$X2)
hometownsplit$X2 <- ifelse(nchar(hometownsplit$X2) > 3, "NaN", hometownsplit$X2)
trim.leading <- function (x)  sub("^\\s+", "", x)
hometownsplit$X3 <- trim.leading(hometownsplit$X3)
hometownsplit$X3 <- ifelse(hometownsplit$X2 %in% hometownsplit$X3, "NR", hometownsplit$X3)
hometownsplit$X3 <- ifelse(hometownsplit$X3 %in% c("NR"), "USA", hometownsplit$X3)

# split the pace column
pacesplit <- data.frame(do.call("rbind", strsplit(as.character(cb_race$Pace), ":", fixed = TRUE)))

# split the time column
timesplit <- data.frame(do.call("rbind", strsplit(as.character(cb_race$Time), ":", fixed = TRUE)))

# rejoin the splits with the data
cb_race <- cbind(cb_race, 
                 TimeHrs = timesplit$X1, 
                 TimeMin = timesplit$X2, 
                 TimeSec = timesplit$X3, 
                 PaceMin = pacesplit$X1, 
                 PaceSec = pacesplit$X2,
                 City = hometownsplit$X1,
                 State = hometownsplit$X2,
                 Country = hometownsplit$X3)


# Format type
cb_race$TimeHrs <- as.numeric(cb_race$TimeHrs)
cb_race$TimeMin <- as.numeric(cb_race$TimeMin)
cb_race$TimeSec <- as.numeric(cb_race$TimeSec)
cb_race$PaceMin <- as.numeric(cb_race$PaceMin)
cb_race$PaceSec <- as.numeric(cb_race$PaceSec)

# Calculate total race time in seconds
cb_race$TotalTimeSec <- round(((cb_race$TimeHrs*60*60) + (cb_race$TimeMin*60) + cb_race$TimeSec), digits = 3)
cb_race$TotalTimeMin <- round(cb_race$TotalTimeSec/60, digits = 3)


# Calculate pace in seconds
cb_race$PaceTimeSec <- round(((cb_race$PaceMin*60) + cb_race$PaceSec), digits = 3)
cb_race$PaceTimeMin <- round(cb_race$PaceTimeSec/60, digits = 3)

# Add age group column
cb_race <- cb_race %>% mutate(AgeGroup = case_when(Age >= 80  & Age <= 89 ~ '80s',
                                             Age >= 70  & Age <= 79 ~ '70s',
                                             Age >= 60  & Age <= 69 ~ '60s',
                                             Age >= 50  & Age <= 59 ~ '50s',
                                             Age >= 40  & Age <= 49 ~ '40s',
                                             Age >= 30  & Age <= 39 ~ '30s',
                                             Age >= 20  & Age <= 29 ~ '20s',
                                             Age >= 10  & Age <= 19 ~ '10s')) # end function

# convert back to cb_race from lists
is.data.frame(cb_race)
is.list(cb_race)
cb_race <- as.list(cb_race)
(y <- do.call(cbind, cb_race))
cb_race <- as.data.frame(y)

# Drop temp columns
drops <- c("TimeHrs", "TimeMin", "TimeSec", "PaceMin", "PaceSec", "Hometown", "V2")
cb_race <- cb_race[,!(names(cb_race) %in% drops)]
colnames(cb_race)[1] <- "Year"

# Set column types
str(cb_race)
cb_race$Year <- as.factor(cb_race$Year)
cb_race$Age <- as.numeric(cb_race$Age)
cb_race$Division <- as.factor(cb_race$Division)
cb_race$TotalTimeMin <- as.numeric(cb_race$TotalTimeMin)
cb_race$TotalTimeSec <- as.numeric(cb_race$TotalTimeSec)
cb_race$PaceTimeMin <- as.numeric(cb_race$PaceTimeMin)
cb_race$PaceTimeSec <- as.numeric(cb_race$PaceTimeSec)
cb_race$AgeGroup <- as.factor(cb_race$AgeGroup)
str(cb_race)

# Convert missing to Na's
cb_race[cb_race == "NaN" ] <- NA
cb_race[cb_race == "NR" ] <- NA

# save out rda file
write.table(cb_race, "../QTW_Spring2021/Unit4_CaseStudy/data/CB_Edited_v2.Rda")
```

# Data Loading:

## Read in clean data
```{r read in RDA - start here for clean data set}
cb_race <- read.table("../QTW_Spring2021/Unit4_CaseStudy/data/CB_Edited_v2.Rda")
```

# Business Analysis

## Exploratory Data Analysis
```{r EDA}
#number of nulls
sum(is.na(cb_race$Age)) ## 20 
nulls <- subset(cb_race, is.na(Age))

#Graph of nulls by year
nulls %>%
 ggplot() +
 aes(x = Year) +
 geom_histogram(bins = 30, fill = "#FC4A9B") +
 theme(plot.title = element_text(hjust=0.5)) +
 labs(x="Year", y="Number of No Age", title = "Number of Runners with No Age Record") +
 theme_minimal()

#Remove nulls from cb_race
cb_race <- na.omit(cb_race)

#Count of runners by Year
cb_race %>%
 filter(!(Division %in% "")) %>%
 filter(!(Country %in% "")) %>%
 ggplot() +
 aes(x = Year) +
 geom_histogram(bins = 30, fill = "#FC4A9B") +
 theme(plot.title = element_text(hjust=0.5)) +
 labs(x="Year", y="Number of Runners", title = "Total Runners Over Time") +
 theme_minimal()

#Avg growth of runners per year = 12% 
YearCnt <- cb_race %>% group_by(Year) %>% summarise(n=n())

#Country group by 
CountryCnt <- cb_race %>% group_by(Country) %>% summarise(n=n())

#Records have null as Country/City/State
  #cb_race[cb_race$State=='mo',]
  #[cb_race$Country=="NY",] 

#Fixing the rows that have incorrect Country, to be listed under state
cb_race[cb_race$State=='va' ,"State"] <- "VA"
cb_race[cb_race$State=='ga' ,"State"] <- "GA"
cb_race[cb_race$State=='mo' ,"State"] <- "MO"

#Top USA States in last 5 years
ByState <- cb_race %>% group_by(State) %>%dplyr::summarise(StateCount = n())
ByState <- ByState[order(-ByState$StateCount),] 

#Top 3 states make up 84% of all runners:
sum(select(top_n(ByState, 3), 'StateCount')) / sum(ByState$StateCount)

#Subsetting the data into top/bottoms
Top10 = subset(ByState[1:10,])
Bottom10 = subset(ByState[46:55,])

#Top 10 Graph
ggplot(Top10, aes(x=reorder(State,-StateCount),
                  y=StateCount,fill=State)) + 
                  geom_bar(stat = "identity") + 
                  theme_minimal() + 
                  theme(plot.title = element_text(hjust=0.5)) + 
                  labs(x="State", y="Number of Runners", title = "Top 10 States With The Most Runners")

#Count of AgeGroup by Year
AgeCount <- cb_race %>% group_by(AgeGroup,Year) %>% summarise(n=n())
#The 20's & 30's group makes up about 75.5 percent of total runners in 2012

#Average Time
cb_race12 <- subset(cb_race, cb_race$Year==2012)
mean(times(cb_race12$Total.Time))
#01:39:01

#Does age need cleaning?
#AgeClean <- cb_race %>% group_by(Age) %>% summarise(n=n())

# Plot of total race time by age group
cb_race %>%
 filter(!is.na(AgeGroup)) %>%
 ggplot() +
 aes(x = "", y = TotalTimeMin, fill = AgeGroup) +
 geom_boxplot() +
 scale_fill_hue() +
 labs(y = "Total Race Time (Minutes)", title = "Womens 10M Cherry Blossom Race", subtitle = "Total Race Time by Age Group") +
 theme_bw()

# Plot of ages by division
ggplot(cb_race) +
 aes(x = "", y = Age, fill = Division) +
 geom_boxplot() +
 scale_fill_hue() +
 labs(y = "Age", title = "Womens 10M Cherry Blossom Race", subtitle = "Total Race Time by Age Group") +
 theme_bw()

# Completion time by division
cb_race %>%
 filter(!is.na(Division)) %>%
 ggplot() +
 aes(x = "", y = TotalTimeMin, fill = Division) +
 geom_boxplot() +
 scale_fill_hue() +
 labs(y = "Total Race Time (Minutes)", title = "Womens 10M Cherry Blossom Race", subtitle = "Total Race Time by Age Division") +
 theme_bw()

# Total time by age group over time
cb_race %>%
 filter(!is.na(AgeGroup)) %>%
 ggplot() +
 aes(x = Year, y = TotalTimeMin, colour = AgeGroup) +
 geom_boxplot(fill = "#0c4c8a") +
 labs(y = "Total Race Time (Minutes)", title = "Womens 10M Cherry Blossom Race", subtitle = "Total Race Time by Age Group Over Time") +
 scale_color_hue() +
 theme_bw()

#plot avg runtime by AgeGroup
cb_race %>% group_by(Year = as.integer(Year)+1998, AgeGroup) %>% 
  summarise(AvgRunTime = mean(TotalTimeMin)) %>%
  ggplot() + aes(x = Year, y = AvgRunTime, colour = AgeGroup) + geom_line() +
 scale_color_hue() + theme_bw() + scale_x_continuous(breaks=c(1999:2012))

#Caclulate percent change each year
cb_race %>% group_by(Year = as.integer(Year)+1998) %>% 
  summarise(AvgRunTime = mean(TotalTimeMin)) %>%
mutate(pct_change = (AvgRunTime/lag(AvgRunTime) - 1) * 100)


```

## Age distribution of runners from 1999-2012
```{r}
# runner age over each year
cb_race$Year <- as.factor(cb_race$Year)
ggplot(cb_race) +
 aes(x = Year, y = Age, fill = Year) +
 geom_boxplot() +
 scale_fill_hue() +
 labs(x = "Year of Race", y = "Age of Runner", title = "Cherry Blossom Race", subtitle = "Runner Age over Years") +
 theme_bw()

# get mean age at each year
age_over_years <- aggregate(cb_race$Age, list(cb_race$Year), mean)
colnames(age_over_years)[1] <- "Year"
colnames(age_over_years)[2] <- "MeanAge"
age_over_years$MeanAge <- round(age_over_years$MeanAge, digits = 3)
age_over_years

# calculate % change year over year
perc_chg <- setDT(age_over_years)[, new.col := MeanAge/shift(MeanAge) - 1]
colnames(perc_chg)[3] <- "PerChng"
perc_chg$PerChng <- round(perc_chg$PerChng, digits = 3)*100
perc_chg

# average change over all years combined
mean_perc_chg <- subset(perc_chg, perc_chg$Year != 1999)
mean_perc_chg$PerChng <- as.numeric(mean_perc_chg$PerChng)
mean_perc_chg <- mean(mean_perc_chg$PerChng)
mean_perc_chg

# plot of average age over years
ggplot(perc_chg) +
 aes(x = Year, y = MeanAge) +
 geom_line(size = 1, group = "#482878") +
 labs(y = "Average Runner Age", title = "Cherry Blossom Race", subtitle = "Average Runner Age Over Time") +
 theme_bw()

# density plot 1
ggplot(cb_race) +
 aes(x = Age, fill = Year) +
 geom_density(adjust = 1L) +
 scale_fill_hue() +
 labs(title = "Cherry Blossom Race", subtitle = "Runner Age Density Plots") +
 theme_bw() +
 facet_grid(vars(Year), vars())

# density plot 2
ggplot(cb_race) +
 aes(x = Age, fill = Year) +
 geom_density(adjust = 1L, alpha = 0.5) +
 scale_fill_hue() +
 labs(title = "Cherry Blossom Race", subtitle = "Density Plot of Runner Age over Time") +
 theme_bw()

```

## Racer Spotlight
```{r}
dupe = cb_race[,c("Name","City", "State")] # select columns to check duplicates
dupe2 = cb_race[duplicated(dupe) | duplicated(dupe, fromLast=TRUE),]
dupe3 = as.data.frame((table(dupe2$Name)))

# Angela Clark, Ann Robb, Betty Blank, Blake Rushin, Kate Mount, Lourdes Hinayon, Malissa Zimmerman, Muffet Chatterton, Sunny Fitzgerald have all ran the race more 13 more times.

spotlight <- cb_race[ which(Name=="Muffet Chatterton"),]

```

```{r age_plot, eval=FALSE}
# Build transparent plots for PPT:
# Mean age by year
p <- ggplot(perc_chg) +
 aes(x = Year, y = MeanAge) +
 geom_line(size = 1, group = "#482878") +
 labs(y = "Average Runner Age", x='') +
# theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA)
  )
p
ggsave(plot = p, file = "../QTW_Spring2021/Unit4_CaseStudy/fig/age.png", 
       bg = "transparent",
       width = 14, height = 5, units = "cm", dpi = 800)


#Age Distribution
p <- ggplot(cb_race) +
  aes(x = Year, y = Age) +
 geom_boxplot(fill = 'hotpink1', color='black') +
 scale_fill_hue() +
 labs(x = "", y = "Age") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
p


ggsave(plot = p, file = "../QTW_Spring2021/Unit4_CaseStudy/fig/age_dist.png", 
       bg = "transparent",
       width = 12, height = 8, units = "cm", dpi = 800)   


# Completion time distribution by division
p <- cb_race %>%
 filter(!is.na(Division)) %>%
 ggplot() +
 aes(x = Division, y = TotalTimeMin) +
 geom_violin(fill='hotpink1', color='black') +
# scale_fill_hue() +
 labs(y = "Race Time (Minutes)") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none",
        axis.text.x = element_text(angle = 45))

p

ggsave(plot = p, file = "../QTW_Spring2021/Unit4_CaseStudy/fig/time_dist.png", 
       bg = "transparent",
       width = 14, height = 5, units = "cm", dpi = 800) 

#Division Distribution


p <- cb_race %>%
 filter(!is.na(Division)) %>%
  filter(Year == 2012) %>%
  group_by(Division) %>% summarise(Runners = n()) %>%
 ggplot() +
 aes(x = Division, y = Runners) +
 geom_col(fill='hotpink1', color='black') +
# scale_fill_hue() +
 labs(y = "Runners") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none",
        axis.text.x = element_text(angle = 45))
p

ggsave(plot = p, file = "../QTW_Spring2021/Unit4_CaseStudy/fig/division_dist_v2.png", 
       bg = "transparent",
       width = 14, height = 5, units = "cm", dpi = 800) 


# simple linear regression for expected number of runners for next year:
runnersPerYear <- cb_race %>%
 filter(!is.na(Division)) %>%
  group_by(Year) %>% summarise(Runners = n())

runnersPerYear$Year = as.integer(runnersPerYear$Year)+1998
fit <- lm(Runners~Year, data=runnersPerYear)
predict(fit, data.frame(Year=2013)) #9,998

summary(fit)
runnersPerYear

```

## ANOVA

There is significant results that the Total time it takes a woman to run the race is different by Age group over the years. 

```{r anova}
attach(cb_race)
mysummary<-function(x){
  result<-c(length(x),mean(x),sd(x),sd(x)/length(x))
  names(result)<-c("N","Mean","SD","SE")
  return(result)
}
sumstats<-aggregate(TotalTimeMin~Year*AgeGroup,data=cb_race,mysummary)
sumstats<-cbind(sumstats[,1:2],sumstats[,-(1:2)])
sumstats


#Standard Deviation plot
ggplot(sumstats,aes(x=Year,y=Mean,group=AgeGroup,colour=AgeGroup))+
  ylab("Total Time in Minutes")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD),width=.1)

#Based on this std dev graph, it looks like there is a clear distinction between means by age group... we might possible need to do multiple ANOVAS to compare different ages across the years. 


#PAUL's
# Compute the analysis of variance
res.aov <- aov(TotalTimeMin ~ AgeGroup, data = cb_race)
#res.aov <- aov(TotalTimeMin ~ Division, data = cb_race)

# Summary of the analysis
summary(res.aov)

# p-value of 2e-16 indicates that there is strong evidence to suggest that the mean race time in minutes by age group is different.

# Difference between groups
diffs <- TukeyHSD(res.aov)
# The difference between 20s & 10s, 80's & 60's, 80's & 70's age groups indicate that there is no evidence for a difference between the mean finish times of those racers.
diffs


# Plot residual diagnostics:
par(mfrow=c(2,2))
plot(res.aov)


par(mfrow=c(1,2))
res = residuals(res.aov)
hist(res)
plot(density(res))

par(mfrow=c(1,1))
```

## ANOVA (cont)
Currently we started to look at the two largest age groups of runners in the race: women in their 20s and 30s. 

```{r ANOVA by 20s}
cb_race20s <- subset(cb_race, cb_race$AgeGroup=="20s")

# For those women in their 20s did their avg time vary across the years
res.aov2 <- aov(TotalTimeMin ~ Year, data = cb_race20s)
# Summary of the analysis
summary(res.aov2)

sumstats2<-aggregate(TotalTimeMin~Year*AgeGroup,data=cb_race20s,mysummary)
sumstats2<-cbind(sumstats2[,1:2],sumstats2[,-(1:2)])
sumstats2

ggplot(sumstats2,aes(x=Year,y=Mean,group=AgeGroup,colour=AgeGroup))+
  ggtitle("Age Variance of Women in their 20s over time") +
  ylab("Total Time in Minutes")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD),width=.1)

# race times have increased in the 20s group since 1999

```

```{r ANOVA by 30s}
cb_race30s <- subset(cb_race, cb_race$AgeGroup=="30s")

# For those women in their 20s did their avg time vary across the years
res.aov3 <- aov(TotalTimeMin ~ Year, data = cb_race30s)
# Summary of the analysis
summary(res.aov3)

sumstats3<-aggregate(TotalTimeMin~Year*AgeGroup,data=cb_race30s,mysummary)
sumstats3<-cbind(sumstats3[,1:2],sumstats3[,-(1:2)])
sumstats3

ggplot(sumstats3,aes(x=Year,y=Mean,group=AgeGroup,colour=AgeGroup))+
  ggtitle("Age Variance of Women in their 30s over time") +
  ylab("Total Time in Minutes")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD),width=.1)

# race times have increased in the 30s group since 1999

```

```{r LOESS}
#plot avg runtime by AgeGroup
cb_race %>% group_by(Year = as.integer(Year)+1998, AgeGroup) %>% 
  summarise(AvgRunTime = mean(TotalTimeMin)) %>%
  ggplot() + aes(x = Year, y = AvgRunTime, colour = AgeGroup) + geom_smooth() +
 scale_color_hue() + theme_bw() + scale_x_continuous(breaks=c(1999:2012))

```















